Switch equivalence and weak prime domains for left-linear DPO rewriting systems

### 2024-12-03 full_version/permutations.tex
fine permutazioni

@@

\begin{definition}
Let $(\X, \R)$ be a left-linear DPO rewriting system. The category $\ppi$ is defined as follows: 
	\begin{itemize}
	\item objects are isomorphism classes of objects of $\X$;
	\item an arrow $[G]\to [H]$ is an equivalence class $[\der{D}, \alpha, \omega]_p$ of abstract decorated derivations between $[G]$ and $[H]$;
	\item given their composition is defined as
	\[[\der{D}_2, \alpha_2, \omega_2]_p \circ [\der{D}_1, \alpha_1, \omega_1]_p:=[(\der{D}_1, \alpha_1, \omega_1) \cdot (\der{D}_2, \alpha_2, \omega_2)]_p\] 
	\item the identity on $[G]$ is $[G, \alpha, \alpha]_p$, where $\alpha$ is any isomorphism $\pi(G)\to G$.	\end{itemize}
\end{definition}
\begin{remark}By \Cref{rem:empty2}
\end{remark}

We are now going to proceed towards the central result of this section: the uniqueness of a consistency permutation between two decorated derivations in a consuming rewriting system.

@@

\begin{lemma}[Restriction Lemma]\label{lem:presuffix} Let $(\X, \R)$ be a consuming left-linear DPO-rewriting system.  Let also $(\der{D}_1, \alpha_1, \omega_1)$, $(\der{D}_2, \alpha_2, \omega_2)$ be two composable decorated derivations of length, respectively, $l_1$ and $l_2$. Suppose that $\sigma\colon [0, l_1+l_2-1]\to [0, l_1+l_2-1]$ is a consistent permutation between $(\der{D}_1, \alpha_1, \omega_1)\cdot (\der{D}_2, \alpha_2, \omega_2)$ and $(\der{D}'_1, \alpha'_1, \omega'_1)\cdot (\der{D}'_2, \alpha'_2, \omega'_2)$ for some other two composable decorated derivations $(\der{D}_1, \alpha_1, \omega_1)$ and $(\der{D}_2, \alpha_2, \omega_2)$.  If $\tau\colon [0,l_1-1]\to [0, l_1-1]$ is a consistent permutation between $(\der{D_1}, \alpha_1, \omega_1)$ and $(\der{D}'_1, \alpha'_1, \omega'_1)$. Then the following hold true:
	\begin{enumerate}
		\item for every $i\in [0, l_2]$ there exist a unique $\zeta_i\colon G_{2,i}\to \lpro \der{D}'_2\rpro $ and a unique $\varsigma_{i}\colon G'_{2,i}\to \lpro \der{D}_2\rpro$ fitting in the diagrams below;
		\[\xymatrix@C=30pt{G_{2,i} \ar[r]^-{\iota_{2,G_{2,i}}} \ar@{.>}[d]_{\zeta_i}&\lpro \der{D}_2 \rpro \ar@{>->}[r]^{q_2}& \tpro{D} \ar[d]^{\xi_{\sigma}} &G'_{2,i} \ar[r]^-{\iota'_{2,G'_{2,i}}} \ar@{.>}[d]_{\varsigma_i}&\lpro \der{D}'_2 \rpro \ar@{>->}[r]^{q'_2}& \lpro D'\rpro \ar[d]^{\xi^{-1}_{\sigma}}\\
			\lpro \der{D}'_2 \rpro \ar@{>->}[rr]_{q'_2}&& \lpro \der{D}'\rpro & 			\lpro \der{D}_2 \rpro \ar@{>->}[rr]_{q_2}&& \lpro \der{D}\rpro }\]

		\item there exists an isomorphism $\psi\colon \lpro \der{D}_2\rpro \to \lpro \der{D}'_2\rpro$ fitting in the diagrams below; 
		\[\xymatrix@C=30pt{G_{2,i} \ar[r]^-{\iota_{2,G_{2,i}}} \ar@/_.3cm/[dr]_{\zeta_i}&\lpro \der{D}_2 \rpro \ar@{>->}[r]^{q_2} \ar@{.>}[d]_{\psi}& \tpro{D} \ar[d]^{\xi_{\sigma}} &G'_{2,i} \ar[r]^-{\iota'_{2,G'_{2,i}}} \ar@/_.3cm/[dr]_{\varsigma_i}&\lpro \der{D}'_2 \rpro\ar@{<.}[d]_{\psi}  \ar@{>->}[r]^{q'_2}& \lpro \der{D}'\rpro \ar[d]^{\xi^{-1}_{\sigma}}\\&
			\lpro \der{D}'_2 \rpro \ar@{>->}[r]_{q'_2}& \lpro \der{D}'\rpro & 		&	\lpro \der{D}_2 \rpro \ar@{>->}[r]_{q_2}& \lpro \der{D}\rpro }\]
		\item the permutation
		\[\varrho\colon [0,l_2-1]\to [0, l_2-1] \qquad i \mapsto \sigma(i+l_1)-l_1\]
		is a consistent one.
	\end{enumerate}
\end{lemma}
\begin{proof}
	To fix the notation, we will use $\der{D}=\{\dder{D}_i\}_{i=0}^{l_1+l_2-1}$ and $\der{D}'=\{\dder{D}_i\}_{i=0}^{l_1+l_2-1}$ to denote the composite derivations. The direct derivations $\dder{D}_i$ and $\dder{D}_i$ are given by the two diagrams below.
	\[\xymatrix{L_i \ar[d]_{m_i}& K_i \ar[r]^{r_i} \ar@{>->}[l]_{l_i} \ar[d]_{k_i}& R_i \ar[d]^{h_i}& L'_i \ar[d]_{m'_i} & K'_i \ar[d]_{k'_i} \ar[r]^{r'_i} \ar@{>->}[l]_{l'_i} & R'_i \ar[d]^{h'_i}\\ G_i & D_i \ar[r]_{g_i} \ar@{>->}[l]^{f_i} & G_{i+1} & G'_i & D'_i \ar[r]_{g'_i} \ar@{>->}[l]^{f'_i} & G'_{i+1}}\]

	We can also notice that,  by \Cref{lem:impo}, $\sigma(i)=\tau(i)$ for every $i\in  [0, l_1-1]$. In particular, this implies that the image  $[l_1, l_1+l_2-1]$ through $\sigma$ is again the interval $[l_1, l_1+l_2-1]$ and that $\varrho$ is well-defined.  Moreover, if $l_1\neq 0$, then $l_1-1$ is the maximum of $E_{\sigma,\tau}$.

@@

		\smallskip \noindent $i=0$. We know that $G_{2,0}$ and $G'_{2,0}$ are isomorphic to, respectively, $G_{1, l_1}$ and $G'_{1, l_1}$. By hypothesis $\tau$ is a consistent permutation between $(\der{D}_1, \alpha_1, \omega_1)$ and $(\der{D}'_1, \alpha'_1, \omega'_1)$, so that $G_{1,l_1}$ and $G'_{1, l_1}$ are isomorphic too. We can therefore consider as $\zeta_0$ the arrow $\iota'_{2, G'_{2,0}}\circ (\alpha'_2)\circ \alpha^{-1}_2$. Now, by \Cref{rem:dett,prop:uniqu}, all subdiagrams of the following diagram commute, so that the whole diagram is commutative and the thesis follows.
		\[\xymatrix@R=10pt@C=35pt{G_{2,0}\ar[r]^{\iota_{2, G_{2,0}}} \ar[dd]_{\alpha^{-1}_2}&\lpro \der{D}_2\rpro \ar[rr]^{q_2}&&\tpro{D}\ar[dddd]^{\xi_{\sigma}}\\& G_{1,l_1} \ar[r]^{\iota_{1, G_{1, l_1}}}& \lpro \der{D}_1\ar[dd]^{\xi_\tau} \ar[ur]_{q_1} \rpro \\\pi(G_{2,0}) \ar[ur]^{\omega_1}\ar[dr]_{\omega'_1} \\ & G'_{1,l_1} \ar[r]_{\iota'_{1, G'_{1, l_1}}}& \lpro \der{D}'_1 \rpro \ar[dr]^{q_1}\\G'_{2,0} \ar@{<-}[uu]^{\alpha'_2}\ar[r]_{\iota'_{2, G'_{2,0}}}&\lpro \der{D}'_2\rpro \ar[rr]_{q'_2}&& \lpro \der{D}'\rpro }\]

@@

\begin{corollary}\label{cor:epi}
	contenuto...
\end{corollary}

>>>>

\begin{remark}\label{rem:ugu}
	It is worth noticing that, by \Cref{rem:empty2}, the permutation equivalence relation reduces to abstract equivalence in the case of empty derivations. This means that, given an object $G$, we have ~$[\{G\}, \alpha, \omega]_a=[\{G\}, \alpha, \omega]_p$.
\end{remark}

@@

\begin{definition}
Let $(\X, \R)$ be a left-linear DPO rewriting system. The category $\ppi$ is defined as follows: 
	\begin{itemize}
	\item objects are isomorphism classes of objects of $\X$;
	\item an arrow $[G]\to [H]$ is an equivalence class $[\der{D}, \alpha, \omega]_p$ of abstract decorated derivations between $[G]$ and $[H]$;
	\item given two arrows $[\der{D}_1, \alpha_1, \omega_1] _p\colon [G]\to [H]$ and $[\der{D}_2, \alpha_2, \omega_2] _p\colon [H]\to [K]$, their composition is defined as
	\[[\der{D}_2, \alpha_2, \omega_2]_p \circ [\der{D}_1, \alpha_1, \omega_1]_p:=[(\der{D}_1, \alpha_1, \omega_1) \cdot (\der{D}_2, \alpha_2, \omega_2)]_p\] 
	\item the identity on $[G]$ is $[\{G\}, \alpha, \alpha]_p$, where $\alpha$ is any isomorphism $\pi(G)\to G$.	\end{itemize}
\end{definition}
\begin{remark} We can point out two facts;
	\begin{itemize}
		\item the composition 	$[\der{D}_2, \alpha_2, \omega_2]_p \circ [\der{D}_1, \alpha_1, \omega_1]_p$ coincides with the permuation equivalence class of $	[\der{D}_2, \alpha_2, \omega_2]_a \circ [\der{D}_1, \alpha_1, \omega_1]_a$;
		\item by \Cref{rem:ugu}, $[\{G, \alpha, \alpha\}]_p=[\{G, \alpha, \alpha\}]_a$, guaranteeing that it is the neutral element with respect of the composition we have defined above.
	\end{itemize}

	In particular, there is a quotient functor $\Pi\colon \dpi\to \ppi$ which is the identity on objects and sends an abstraction equivalence class to its permutation equivalence one.
\end{remark}

\subsection{Uniqueness of consistent permutations}

We are now going to proceed towards the central result of this section: the uniqueness of a consistent permutation between two decorated derivations in a consuming DPO rewriting system.

@@

The previous results, in turn, allows us to deduce a property which is in some sense the converse of the Composition \Cref{lem:sum}.

\begin{lemma}[Restriction Lemma]\label{lem:presuffix} Let $(\X, \R)$ be a consuming left-linear DPO-rewriting system.  Let also $(\der{D}_1, \alpha_1, \omega_1)$, $(\der{D}_2, \alpha_2, \omega_2)$ be two composable decorated derivations of length, respectively, $l_1$ and $l_2$ such that there are  other two composable decorated derivations $(\der{D}_1, \alpha_1, \omega_1)$ and $(\der{D}_2, \alpha_2, \omega_2)$ with the property that
	\[[(\der{D}_1, \alpha_1, \omega_1)\cdot (\der{D}_2, \alpha_2, \omega_2)]_a\equiv^p_\sigma[(\der{D}'_1, \alpha'_1, \omega'_1)\cdot (\der{D}'_2, \alpha'_2, \omega'_2)]_a\]

Suppose, finally, that $[\der{D_1}, \alpha_1, \omega_1]_a \equiv^p_\tau [\der{D}'_1, \alpha'_1, \omega'_1]_a$ for some $\tau$. Then the following hold true:
	\begin{enumerate}
		\item for every $i\in [0, l_2]$ there exist a unique $\zeta_i\colon G_{2,i}\to \lpro \der{D}'_2\rpro $ and a unique $\varsigma_{i}\colon G'_{2,i}\to \lpro \der{D}_2\rpro$ fitting in the diagrams below;
		\[\xymatrix@C=30pt{G_{2,i} \ar[r]^-{\iota_{2,G_{2,i}}} \ar@{.>}[d]_{\zeta_i}&\lpro \der{D}_2 \rpro \ar@{>->}[r]^{q_2}& \tpro{D} \ar[d]^{\xi_{\sigma}} &G'_{2,i} \ar[r]^-{\iota'_{2,G'_{2,i}}} \ar@{.>}[d]_{\varsigma_i}&\lpro \der{D}'_2 \rpro \ar@{>->}[r]^{q'_2}& \lpro D'\rpro \ar[d]^{\xi^{-1}_{\sigma}}\\
			\lpro \der{D}'_2 \rpro \ar@{>->}[rr]_{q'_2}&& \lpro \der{D}'\rpro & 			\lpro \der{D}_2 \rpro \ar@{>->}[rr]_{q_2}&& \lpro \der{D}\rpro }\]

		\item there exists an isomorphism $\psi\colon \lpro \der{D}_2\rpro \to \lpro \der{D}'_2\rpro$ fitting in the diagrams below; 
		\[\xymatrix@C=30pt{G_{2,i} \ar[r]^-{\iota_{2,G_{2,i}}} \ar@/_.3cm/[dr]_{\zeta_i}&\lpro \der{D}_2 \rpro \ar@{>->}[r]^{q_2} \ar@{.>}[d]_{\psi}& \tpro{D} \ar[d]^{\xi_{\sigma}} &G'_{2,i} \ar[r]^-{\iota'_{2,G'_{2,i}}} \ar@/_.3cm/[dr]_{\varsigma_i}&\lpro \der{D}'_2 \rpro\ar@{<.}[d]_{\psi}  \ar@{>->}[r]^{q'_2}& \lpro \der{D}'\rpro \ar[d]^{\xi^{-1}_{\sigma}}\\&
			\lpro \der{D}'_2 \rpro \ar@{>->}[r]_{q'_2}& \lpro \der{D}'\rpro & 		&	\lpro \der{D}_2 \rpro \ar@{>->}[r]_{q_2}& \lpro \der{D}\rpro }\]
		\item the permutation
		\[\varrho\colon [0,l_2-1]\to [0, l_2-1] \qquad i \mapsto \sigma(i+l_1)-l_1\]
		is a consistent one.
	\end{enumerate}
\end{lemma}

\begin{proof}
	To fix the notation, we will use $\der{D}=\{\dder{D}_i\}_{i=0}^{l_1+l_2-1}$ and $\der{D}'=\{\dder{D}_i\}_{i=0}^{l_1+l_2-1}$ to denote the composite derivations. The direct derivations $\dder{D}_i$ and $\dder{D}_i$ are given by the two diagrams below.
	\[\xymatrix{L_i \ar[d]_{m_i}& K_i \ar[r]^{r_i} \ar@{>->}[l]_{l_i} \ar[d]_{k_i}& R_i \ar[d]^{h_i}& L'_i \ar[d]_{m'_i} & K'_i \ar[d]_{k'_i} \ar[r]^{r'_i} \ar@{>->}[l]_{l'_i} & R'_i \ar[d]^{h'_i}\\ G_i & D_i \ar[r]_{g_i} \ar@{>->}[l]^{f_i} & G_{i+1} & G'_i & D'_i \ar[r]_{g'_i} \ar@{>->}[l]^{f'_i} & G'_{i+1}}\]

	We can also notice that,  by \Cref{lem:impo}, $\sigma(i)=\tau(i)$ for every $i\in  [0, l_1-1]$. In particular, this implies that the image  $[l_1, l_1+l_2-1]$ through $\sigma$ is again the interval $[l_1, l_1+l_2-1]$, that $\varrho$ is well-defined and $\sigma=\tau +\varrho$.  Moreover, if $l_1\neq 0$, then $l_1-1$ is the maximum of $E_{\sigma,\tau}$.

@@

		\smallskip \noindent $i=0$. We know that $G_{2,0}$ and $G'_{2,0}$ are isomorphic to, respectively, $G_{1, l_1}$ and $G'_{1, l_1}$. By \Cref{prop:abs}, $\tau$ is a consistent permutation between $(\der{D}_1, \alpha_1, \omega_1)$ and $(\der{D}'_1, \alpha'_1, \omega'_1)$, so that $G_{1,l_1}$ and $G'_{1, l_1}$ are isomorphic too. We can therefore consider as $\zeta_0$ the arrow $\iota'_{2, G'_{2,0}}\circ (\alpha'_2)\circ \alpha^{-1}_2$. Now, by \Cref{rem:dett,prop:uniqu}, all subdiagrams of the following diagram commute, so that the whole diagram is commutative and the thesis follows.
		\[\xymatrix@R=10pt@C=35pt{G_{2,0}\ar[r]^{\iota_{2, G_{2,0}}} \ar[dd]_{\alpha^{-1}_2}&\lpro \der{D}_2\rpro \ar[rr]^{q_2}&&\tpro{D}\ar[dddd]^{\xi_{\sigma}}\\& G_{1,l_1} \ar[r]^{\iota_{1, G_{1, l_1}}}& \lpro \der{D}_1\ar[dd]^{\xi_\tau} \ar[ur]_{q_1} \rpro \\\pi(G_{2,0}) \ar[ur]^{\omega_1}\ar[dr]_{\omega'_1} \\ & G'_{1,l_1} \ar[r]_{\iota'_{1, G'_{1, l_1}}}& \lpro \der{D}'_1 \rpro \ar[dr]^{q_1}\\G'_{2,0} \ar@{<-}[uu]^{\alpha'_2}\ar[r]_{\iota'_{2, G'_{2,0}}}&\lpro \der{D}'_2\rpro \ar[rr]_{q'_2}&& \lpro \der{D}'\rpro }\]

@@

\begin{corollary}\label{cor:epi}
	Let $(\X, \R)$ be a consuming left-linear DPO rewriting, then in $\ppi$ every arrow is epi.
\end{corollary}
\begin{proof}
Let $[\der{D}, \alpha, \omega]_p$ be an arrow in $\ppi$ between $[G]\to [H]$ and suppose that two arrows $[\der{D}_1, \alpha_1, \omega_1]_p, [\der{D}_2, \alpha_2, \omega_2]_p\colon [H]\to [K]$  are given such that 
\[[\der{D}_1, \alpha_1, \omega_1]_p\circ [\der{D}, \alpha, \omega]_p=[\der{D}_2, \alpha_2, \omega_2]_p\circ [\der{D}, \alpha, \omega]_p\]

By construction and \Cref{prop:abs}, the previous equation entails the existence of a consistent permutation $\sigma$ between $(\der{D}, \alpha, \omega)\cdot (\der{D}_1, \alpha_1, \omega_1)$ and $(\der{D}, \alpha, \omega)\cdot (\der{D}_2, \alpha_2, \omega_2)$. Since the identity is a consistent permutation between $(\der{D}, \alpha, \omega)$ and itself, the thesis follows immediately from \Cref{lem:presuffix}.
\end{proof}
